---
title: "Splitting a large dataset, running cellpose, and measuring the results using fiji tracker."
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
  html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    keep_md: false
    mode: selfcontained
    number_sections: true
    self_contained: true
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

<style type="text/css">
body, td {
  font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
 font-size: 16px
}
</style>

```{r options, include=FALSE}
knitr::opts_knit$set(
  width = 120, progress = TRUE, verbose = TRUE, echo = TRUE)
knitr::opts_chunk$set(error = TRUE, dpi = 96)
lua_filters <- rmarkdown::pandoc_lua_filter_args("pandoc-zotxt.lua")
old_options <- options(
  digits = 4, stringsAsFactors = FALSE, knitr.duplicate.label = "allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
rundate <- format(Sys.Date(), format = "%Y%m%d")
previous_file <- ""
ver <- format(Sys.Date(), "%Y%m%d")

##tmp <- sm(loadme(filename=paste0(gsub(pattern="\\.Rmd", replace="", x=previous_file), "-v", ver, ".rda.xz")))
rmd_file <- "index_functions_big.Rmd"
library(spatstat.geom)
library(reticulate)
```

# Changelog

* 20230508: Trying this out with a larger dataset, explicitly set wanted_z and wanted_channel for it.
* 20230328: Reimplemented steps as functions, connected them in one set of calls.
* 20230317: Initial cellpose implementation.
* 20230315: Separation of input image into timepoints.
* 20230310: Used geopandas to trace cells over time by position.
* 20230307: Setting up my environment to handle ipynb and python markdown.

# Introduction

Jacques is seeking to segment and follow cells over time.

Steps performed so far:

1.  Python/venv installation of fiji
2.  Figured out some easy interactions between python and actual image data.
3.  Implemented a simple function to find minimum distances between
    putative cells using the extant methods.

# Next steps

1.  Note that the czi images are immediately openable via fiji's bioformat interface.
2.  Load dataset via fiji
3.  Invoke cellpose via the python interface (same as pyimagej)
4.  Save ROIs produced by cellpose, save them to zip output file
  a.  The roi interface in fiji can address these
5.  Currently a macro is invoked which performs a set of measurements
    on every ROI and saves them to a csv.  This creates the primary
    data structure used for all following processes.
6.  Need to create a datastructure which identifies each individual
    cell over time given these ROIs which x/y/time(frame) along with
    the measurements of interest (area/mean/stdev/intensities by
    color).
    a.  Small postprocessing details: the intensity values produced
    must be normalized by cell area.

# Demonstrate the first couple of steps using an actual dataset

Jacques sent me an image acquired from the microscope and I saved it
as 'test_data/raw.tif', he also sent me a cellpose model which I saved
to the 'models/' directory.

Given these as a starting point, let us try to open the image with a
fiji instance and split it into a series of smaller files by
timepoint.

# Implementation notes

I have been primarily using this as a way to refresh my brain on
python and get current on best practices.  Thus, there are decisions I
made in this workbook which do not make sense in any other context:
e.g. why would anyone have the primary data structure be a dictionary
keyed by filename, that is dumb?  I chose to do that to refresh myself
on playing with dictionaries.  By the same token, why would anyone
make a dictionary of dataframes only to turn around and concatenate
them for usage in geopandas, that is crazy?  I did this to get more
comfortable with pandas and get out some of my Rish muscle memory.

With that in mind, if we choose to make this a package, the first
thing that will need to happen is to rework the base datastructure.  I
just want anyone who actually reads this code to know that yes, I am a
nutter, but a nutter for a reason.

# Load necessary python modules

I am hoping to figure out how to load my package and have it handle all the initialization.
It is not clear yet to me if I can import fiji_tracker and get access to the various imports it
performs.  I am reasonably certain this is possible, but that I have not figured
out the correct syntax.

```{python load}
from fiji_tracker.fiji_tracker.tracker import *

from cellpose import models, io
from cellpose.io import *
from collections import defaultdict
import geopandas
import glob
import imagej
from jpype import JArray, JInt
import matplotlib.pyplot as plt
import multiprocessing as mp
import numpy as np
import os
import pandas
from pandas import DataFrame
from pathlib import Path
import scyjava
import seaborn
import shutil

base_dir = Path('/lab/scratch/atb/imaging/mtb_2023').as_posix()
os.chdir(base_dir)
input_file = Path(f"{base_dir}/test_data/Experiment-1568.czi").as_posix()
pandas.set_option('display.max_columns', None)
verbose = True
```

```{python start_fiji}
start_dir = os.getcwd()
mem = '-Xmx128g'
location = 'venv/bin/Fiji.app'
mode = 'interactive'

ij, raw_image, imp = start_fiji(base=start_dir, mem=mem, location=location, mode=mode,
                                input_file=input_file)
```

# Separate slices

Note to self, Jacques' new dataset uses wanted_z == 2, wanted_channel == 3.

```{python run_separate_slices}
raw_dataset, saved_slices, output_directory = separate_slices(input_file, ij, wanted_z = 2,
                                                              wanted_channel = 3)
```

# Invoke cellpose

```{python run_invoke_cellpose}
output_files = invoke_cellpose(output_directory, 'models/CP_20220523_104016')

moved = move_cellpose_output(output_files)
```

# Collapse Z

One possible change is to perform measurements on the sum of Z-stacks
instead of a single slice.  Thus we would sum the cells, create the
ROIs using the single slice grayscale image, then measure the set of
all combined.

```{python run_collapse_z}
output_files = collapse_z(raw_dataset, output_files, ij)
```

# Create Regions of interest from cellpose outputs

In Jacques notebook, it looks like he only extracts ROIs from one of
the cellpose slices.  I am assuming the goal is to extend this across
all images?

There is an important caveat that I missed: imagej comes with a
python2-based scripting language from which it appears some of his
code is coming.  As a result I should look carefully before using it,
and pay close attention to the examples provided here for the most
appropriate ways of interacting with the ROI manager etc:

https://github.com/imagej/pyimagej/blob/main/doc/examples/blob_detection_interactive.py

```{python invoke_slice_to_roi}
slice_measurements = slices_to_roi_measurements(output_files, ij, raw_image, imp,
                                                collapsed = True, stop_after = 5)
```

# Convert the slice measurements to pandas df

slices_to_roi_measurements() returns a dictionary with keys which are
the filenames of each raw tif file.  Each element of that dictionary
is in turn a dictionary containing some information about the files
along with a df of the measurements provided by imagej.

My little geopandas function assumes a single long df with some
columns which tell it which timepoint.  So lets make a quick function
to give that here.  OTOH it may be wiser/better to make some changes
to slices_to_roi_measurements() so that it returns that format df; but
since I am using this as a learning experience to get more comfortable
with python data structures, I will not do it that way.

The number of rows of concatenated should be equivalent to the number
of ROIs and in the same order.  Sadly this is not quite true.

```{python invoke_convert_pandas}
concatenated = convert_slices_to_pandas(slice_measurements)

concatenated.shape[0]
concatenated.head()
```

# Find nearest

```{python invoke_nearest_cells}
nearest, traced_ids, paired, pairwise_distances = nearest_cells_over_time(
    concatenated, x_column = 'X', y_column = 'Y')
```

## Using the ROIs to trace cells over time

nearest_cells_over_time() now returns three data structures which I
hope will provide sufficient information to successfully trace and
test the identities of the 'nearest' cells.  One thing which I think
will help would be to rename the ROIs according to the cell IDs
generated by it.

```{python rename_rois, eval=FALSE}
Overlay = scyjava.jimport('ij.gui.Overlay')
ov = Overlay()
rm = ij.RoiManager.getRoiManager()

new_names = []
for group in nearest:
    elements = nearest[group]
    for element in elements:
        combined = f"{group}_{element}"
        print(f"The new name is {combined}")
        new_names.append(combined)

roi_count = rm.getCount()

roi_zero_idx = roi_count - 1
rm.select(roi_zero_idx)
imp.setT(timepoint)
updated2 = rm.runCommand("Update")
print(f"Finished {timepoint}")
rm.setSelectedIndexes(roi_index)



```

## Get information from a group of cells

As a final step, we should be able to extract and play with the
information from one or more groups of cells.

```{python get_info}
cell_id = 2
cell_idx = nearest[cell_id]
cell_data = concatenated.loc[cell_idx]
len(cell_data)
cell_data = cell_data.reset_index()

scatter = plt.scatter(cell_data['X'], cell_data['Y'])
final_row = cell_data.index.max()
for start_time in range(0, final_row - 1):
    ti_idx = cell_data.index == start_time
    tj_idx = cell_data.index == start_time + 1
    p1x = cell_data[ti_idx].X
    p2x = cell_data[tj_idx].X
    p1y = cell_data[ti_idx].Y
    p2y = cell_data[tj_idx].Y
    x_points = [p1x, p2x]
    y_points = [p1y, p2y]
    plt.plot(x_points, y_points)
finalm1_idx = cell_data.index == final_row - 1
final_idx = cell_data.index == final_row
finalm1_x = cell_data[finalm1_idx].X
final_x = cell_data[final_idx].X
finalm1_y = cell_data[finalm1_idx].Y
final_y = cell_data[final_idx].Y
x_points = [finalm1_x, final_x]
y_points = [finalm1_y, final_y]
plt.plot(x_points, y_points)
plt.show()

seaborn.violinplot(data = cell_data.Area)
plt.show()
```

# Testing out function fragments

## Separate the logic of the above function

slices_to_roi_measurements() does a few things in one for loop:

1.  It creates polygons from the cellpose outputs.
2.  It adds them to the ROI manager
3.  It performs a measurement of them.

There are other tasks which Jacques would like to perform on the various cellpose polygons, thus
we really need to separate the functions which create and save the ROIs from those
which measure them.

In addition, I think we really want to have the raw data open and add the ROIs to it.

```{python roi_updater, eval=FALSE}
updated = create_cellpose_rois(output_files, ij, raw_image, imp, collapsed=True)
```

# Test create_cellpose_rois()

The following block was used to try and get create_cellpose_rois working properly.

It is therefore a bit of a mess.  I think everything in it is now in
fiji_tracker/tracker.py's create_cellpose_rois() and it may safely be deleted.

```{python testme, eval=FALSE}
from fiji_tracker.fiji_tracker.tracker import *

from cellpose import models, io
from cellpose.io import *
from collections import defaultdict
import geopandas
import glob
import imagej
from jpype import JArray, JInt
from math import isnan
import matplotlib.pyplot as plt
import multiprocessing as mp
import numpy as np
import os
import pandas
from pandas import DataFrame
from pathlib import Path
import scyjava
import seaborn
import shutil

base_dir = Path('/lab/scratch/atb/imaging/mtb_2023').as_posix()
os.chdir(base_dir)
input_file = Path(f"{base_dir}/test_data/Experiment-1568.czi").as_posix()
pandas.set_option('display.max_columns', None)
verbose = True

collapsed = False
verbose = True
delete = False
max_frames = 2

start_dir = os.getcwd()
mem = '-Xmx64g'
location = 'venv/bin/Fiji.app'
mode = 'interactive'

ij, raw_image, imp = start_fiji(base=start_dir, mem=mem, location=location, mode=mode,
                                input_file=input_file)
raw_dataset, saved_slices, output_directory = separate_slices(input_file, ij, wanted_z = 2,
                                                              wanted_channel = 3)
output_files = invoke_cellpose(output_directory, 'models/CP_20220523_104016')

## Cheating here to get around my for loop
cellpose_slices = list(output_files.keys())[0]

data_info = {}
for element in range(len(raw_image.dims)):
    name = raw_image.dims[element]
    data_info[name] = raw_image.shape[element]
num_times = data_info['Time'] + 1
num_channels = data_info['Channel']
num_z = data_info['Z']

Overlay = scyjava.jimport('ij.gui.Overlay')

rm.runCommand("Associated", "true")
#rm.runCommand("show All with labels")
slice_directory = ''
print("Starting to iterate over times.")

max_frames = 5
for timepoint in range(1, num_times):
    frame_number = timepoint - 1 ## I used 0-indexed for the frames.
    if (timepoint > max_frames):
        print("Stopping while debugging.")
        break
    print(f"Going to time: {timepoint}")
    imp.setT(timepoint)
    slice_name = f"frame_{frame_number}"
    input_tif = output_files[slice_name]['input_file']
    slice_directory_name = os.path.basename(os.path.dirname(os.path.dirname(input_tif)))
    input_txt = output_files[slice_name]['output_txt']
    input_mask = output_files[slice_name]['output_mask']
    ## The logic for this was taken from:
    ## https://stackoverflow.com/questions/73849418/is-there-any-way-to-switch-imagej-macro-code-to-python3-code
    txt_fh = open(input_txt, 'r')
    roi_stats = defaultdict(list)
    frame_xcoords = []
    frame_ycoords = []
    coords_length = []
    ## Now get the slice for this timepoint from the raw data
    seleced = []
    line_count = 0
    max_lines = 5
    for line in txt_fh:
        line_count = line_count + 1
        if (line_count > max_lines):
            break
        xy = line.rstrip().split(",")
        xy_coords = [int(element) for element in xy if element not in '']
        x_coords = [int(element) for element in xy[::2] if element not in '']
        y_coords = [int(element) for element in xy[1::2] if element not in '']
        xcoords_jint = JArray(JInt)(x_coords)
        ycoords_jint = JArray(JInt)(y_coords)
        polygon_roi_instance = scyjava.jimport('ij.gui.PolygonRoi')
        roi_instance = scyjava.jimport('ij.gui.Roi')
        imported_polygon = polygon_roi_instance(xcoords_jint, ycoords_jint,
                                                    len(x_coords), int(roi_instance.POLYGON))
        imp.setRoi(imported_polygon)
        added = rm.addRoi(imported_polygon)
        ## Get the current number of ROIs, 1 indexed.
        roi_count = rm.getCount()
        roi_zero_idx = roi_count - 1
        rm.select(roi_zero_idx)
        imp.setT(timepoint)
        current_name = rm.getName(roi_zero_idx)
        new_name = f"{frame_number}_{line_count}_{current_name}"
        rm.rename(roi_zero_idx, new_name)
        updated2 = rm.runCommand("Update")
        print(f"Finished {timepoint}")
    txt_fh.close()
imp.show()
roizipname = "testROIs.zip"
roizippath = os.path.join(roizipname)
roi_index = JArray(JInt)(range(0, rm.getCount()))
rm.setSelectedIndexes(roi_index)
rm.runCommand('Save', roizippath)



## Note to self, when selecting ROIs for updating, one must do so one
## at a time, lest the window selection be lost and thus not be able
## to perform the update and save the ROI set.
## Similarly one may not use Select All for measurement
## but you can do 'Select' '0,1,2,3...n' followed by 'Measure'.
```


# Test nearest_cells_over_time()

In a similar fashion, Jacques showed that there is a problem with how
sjoin_nearest() is defining the closest cells with respect to time.  I
am therefore putting the body of that function into the following
block in order to figure out what is going wrong and ideally fix it.

In a similar fashion, I think the following block is also no longer needed.

```{python test_nearest_cells_over_time, eval=FALSE}
df = concatenated
max_dist = 80.0
x_column = 'X'
y_column = 'Y'

## One idea which might be interesting: add frame/Z here
gdf = geopandas.GeoDataFrame(
    df, geometry=geopandas.points_from_xy(df[x_column], df[y_column]))
gdf.head()  ## We start with frame 0
gdf.tail()  ## and end with 120.

final_time = gdf.Frame.max()
pairwise_distances = []
for start_time in range(0, final_time):
    i = start_time
    j = i + 1
    ti_idx = gdf.Frame == i
    tj_idx = gdf.Frame == j
    if verbose:
        print(f"Getting distances of dfs {i} and {j}.")
    ti = gdf[ti_idx]
    tj = gdf[tj_idx]
    ti_rows = ti.shape[0]
    tj_rows = tj.shape[0]
    for ti_row in range(0, ti_rows):
        ti_element = ti.iloc[[ti_row, ]]
        titj = geopandas.sjoin_nearest(ti_element, tj, distance_col="pairwise_dist",
                                       max_distance=max_dist)
        chosen_closest_dist = titj.pairwise_dist.min()
        if (isnan(chosen_closest_dist)):
            print(f"This element has no neighbor within {max_dist}.")
        else:
            chosen_closest_cell = titj.pairwise_dist == chosen_closest_dist
            chosen_closest_row = titj[chosen_closest_cell]
            pairwise_distances.append(chosen_closest_row)

paired = pandas.concat(pairwise_distances)
id_counter = 0
## Cell IDs pointing to a list of cells
traced = {}
## Endpoints pointing to the cell IDs
ends = {}
for i in range(0, final_time - 1):
    query_idx = paired.Frame_left == i
    query = paired[query_idx]
    for row in query.itertuples():
        start_cell = row.Index
        end_cell = row.index_right
        if start_cell in ends.keys():
            cell_id = ends[start_cell]
            current_value = traced[cell_id]
            current_value.append(end_cell)
            traced[cell_id] = current_value
            ends[end_cell] = cell_id
        else:
            id_counter = id_counter + 1
            traced[id_counter] = [start_cell, end_cell]
            ends[end_cell] = id_counter
## return traced paired pairwise_distances
```




```{r saveme}
pander::pander(sessionInfo())
message(paste0("This is hpgltools commit: ", get_git_commit()))
```
